---
title: "Econ 103: Project 1"
author: "Mason Lee, Alexander Prater, Fahad Malik, Kalani Leibow"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```


## Packages

```{r}
#install.packages("Lahman")
library(Lahman)
library(tidyverse)
library(ggplot2)
library(car)

```


## Data Loading

```{r}
data(Batting)
head(Batting)
```

## Data Manipulation

```{r}
BattingData <- Batting

BattingData <- BattingData %>%
  filter(yearID >= 1985, AB >= 100, IBB > 0, HR > 0, SO > 0, RBI > 0) %>%
  mutate(BA = H / AB, Singles = (H - X2B - X3B - HR), 
    SLG = ((H - X2B - X3B - HR) + 2 * X2B + 3 * X3B + 4 * HR) / AB,
         OBP = (H + BB + HBP) / (AB + BB + SF + HBP)) %>%
  mutate(OPS = SLG + OBP)

BattingDataSlim <- BattingData[c(1, 2, 5, 12, 13, 17, 18, 25)]

head(BattingDataSlim)

```

## Graphs of 4 Predictors (Question 2)

### 5 Number Summary

```{r}
fivenum(BattingDataSlim$HR)
fivenum(BattingDataSlim$RBI)
fivenum(BattingDataSlim$SLG)
fivenum(BattingDataSlim$SO)
fivenum(BattingDataSlim$IBB)
```


### Histograms

```{r}
hist(BattingDataSlim$HR, breaks = "scott")
hist(BattingDataSlim$RBI, breaks = "scott")
hist(BattingDataSlim$SLG, breaks = "scott")
hist(BattingDataSlim$SO, breaks = "scott")
hist(BattingDataSlim$IBB, breaks = "scott")

```
### Boxplots

```{r}
boxplot(BattingDataSlim$HR)
boxplot(BattingDataSlim$RBI)
boxplot(BattingDataSlim$SLG)
boxplot(BattingDataSlim$SO)
boxplot(BattingDataSlim$IBB)


```


### QQ Plots

```{r}
qqplot(BattingDataSlim$HR, BattingDataSlim$IBB)
qqplot(BattingDataSlim$RBI, BattingDataSlim$IBB)
qqplot(BattingDataSlim$SLG, BattingDataSlim$IBB)
qqplot(BattingDataSlim$SO, BattingDataSlim$IBB)
```

### Scatter Plots with Regression Lines (Question 3)

```{r}

ggplot(BattingDataSlim, aes(x = HR, y = IBB)) +
  geom_point() + 
  geom_jitter() 


ggplot(BattingDataSlim, aes(x = SLG, y = IBB)) +
  geom_point()

ggplot(BattingDataSlim, aes(x = SO, y = IBB)) +
  geom_point()

ggplot(BattingDataSlim, aes(x = RBI, y = IBB)) +
  geom_point() 


```

## Checking for Transformations (Question 4)

```{r}
summary(lm(BattingDataSlim$IBB ~ (BattingDataSlim$HR))) #R^2 0.2063
summary(lm(BattingDataSlim$IBB ~ exp(BattingDataSlim$HR))) #R^2 0.005221
summary(lm(BattingDataSlim$IBB ~ sqrt(BattingDataSlim$HR))) #R^2 0.1718
summary(lm(BattingDataSlim$IBB ~ log(BattingDataSlim$HR))) #R^2 0.1238
```

```{r}
summary(lm(BattingDataSlim$IBB ~ (BattingDataSlim$RBI))) #R^2 0.2305
summary(lm(BattingDataSlim$IBB ~ exp(BattingDataSlim$RBI))) #R^2 5.118e-05 
summary(lm(BattingDataSlim$IBB ~ sqrt(BattingDataSlim$RBI))) #R^2 0.2057
summary(lm(BattingDataSlim$IBB ~ log(BattingDataSlim$RBI))) #R^2 0.1707
```

```{r}
summary(lm(BattingDataSlim$IBB ~ (BattingDataSlim$SLG))) #R^2 0.1487
summary(lm(BattingDataSlim$IBB ~ exp(BattingDataSlim$SLG))) #R^2 0.1601
summary(lm(BattingDataSlim$IBB ~ sqrt(BattingDataSlim$SLG))) #R^2 0.1365
summary(lm(BattingDataSlim$IBB ~ log(BattingDataSlim$SLG))) #R^2 0.1244
```

```{r}
summary(lm(BattingDataSlim$IBB ~ (BattingDataSlim$SO))) #R^2 0.04826
summary(lm(BattingDataSlim$IBB ~ exp(BattingDataSlim$SO))) #R^2 -8.603e-05 
summary(lm(BattingDataSlim$IBB ~ sqrt(BattingDataSlim$SO))) #R^2 0.05132 
summary(lm(BattingDataSlim$IBB ~ log(BattingDataSlim$SO))) #R^2 0.05137 
```

```{r}
densityPlot(BattingDataSlim$HR) #best R^2, will use this for the Transformation
densityPlot(sqrt(BattingDataSlim$HR)) #best visually
densityPlot(log(BattingDataSlim$HR))
densityPlot(exp(BattingDataSlim$HR))

densityPlot(BattingDataSlim$SO) #best visually
densityPlot(sqrt(BattingDataSlim$SO)) #best R^2, will use this for the Transformation
densityPlot(log(BattingDataSlim$SO))
densityPlot(exp(BattingDataSlim$SO))

densityPlot(BattingDataSlim$SLG) #best visually
densityPlot(sqrt(BattingDataSlim$SLG)) 
densityPlot(log(BattingDataSlim$SLG))
densityPlot(exp(BattingDataSlim$SLG)) #best R^2, will use this for the Transformation

densityPlot(BattingDataSlim$RBI) #best visually, best R^2, will use this for the Transformation
densityPlot(sqrt(BattingDataSlim$RBI)) 
densityPlot(log(BattingDataSlim$RBI))
densityPlot(exp(BattingDataSlim$RBI))
```

## Post-Transformation Linear Models (Question 6)

```{r}
model1 <- lm(BattingDataSlim$IBB ~ (BattingDataSlim$HR))
model2 <- lm(BattingDataSlim$IBB ~ sqrt(BattingDataSlim$SO))
model3 <- lm(BattingDataSlim$IBB ~ exp(BattingDataSlim$SLG))
model4 <- lm(BattingDataSlim$IBB ~ (BattingDataSlim$RBI))


summary(model1)
summary(model2)
summary(model3)
summary(model4)

confint(model1)
confint(model2)
confint(model3)
confint(model4)

```

## Bootstrapping (Question 7)

```{r}

model4boot <- Boot(model4, R=1000)

summary(model4boot)
confint(model4boot)
hist(model4boot)

```


```{r}
regression_model <- parsnip::linear_reg() %>%
  parsnip::set_engine("lm") %>%
  parsnip::set_mode("regression")

fit <- regression_model %>%
  parsnip::fit(data = BattingDataSlim, IBB ~ RBI)

summary(fit$fit)
```

```{r}
resid <- broom::augment(fit, BattingDataSlim)

ggplot(data = resid, aes(x = .pred, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red")

```

```{r}
qqnorm(residuals(model4))
qqline(residuals(model4), lwd = 1)
```


# Analysis of Code

#### The following section answers each of the questions explicitly, while also providing commentary on the important graphs