---
title: "HW5 - Econ 144"
author: "Mason Lee"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache  = TRUE)
```

## Packages

```{r packages}
library(readr)
library(lattice)
library(foreign)
library(MASS)
library(car)
require(stats)
require(stats4)
library(KernSmooth)
library(fastICA)
library(cluster)
library(leaps)
library(mgcv)
library(rpart)
library(pan)
library(mgcv)
library(DAAG)
library("TTR")
library(tis)
require("datasets")
require(graphics)
library(forecast)
require(astsa)
library(RColorBrewer)
library(plotrix)
library(tsibble)
library(ggplot2)
library(tidyr)
library(USgas)
library(fpp3)
```


# Textbook A: Forecasting for Economics and Business

## Data

```{r data}
fm <- read.csv("fmhpi_master_file.csv")
head(fm)
table(fm$GEO_Type)
msa <- subset(fm, GEO_Type == "CBSA")
MSA1_name <- "Oxnard-Thousand Oaks-Ventura CA"
MSA2_name <- "Los Angeles-Long Beach-Anaheim CA"

msa1 <- subset(msa, GEO_Name == MSA1_name)
msa2 <- subset(msa, GEO_Name == MSA2_name)

msa1$Date <- as.Date(paste(msa1$Year, msa1$Month, "01", sep = "-"))
msa2$Date <- as.Date(paste(msa2$Year, msa2$Month, "01", sep = "-"))

msa1 <- msa1[order(msa1$Date), ]
msa2 <- msa2[order(msa2$Date), ]

head(msa1)
head(msa2)
```

## 11.1 

```{r 11.1}
dat <- merge(
  data.frame(
    Date = msa1$Date,
    g1   = c(NA, 100 * diff(log(msa1$Index_SA)))  # Oxnard growth
  ),
  data.frame(
    Date = msa2$Date,
    g2   = c(NA, 100 * diff(log(msa2$Index_SA)))  # LA growth
  ),
  by = "Date"
)
dat <- dat[complete.cases(dat), ]

summary(dat)
cor(dat$g1, dat$g2)

dat$Month <- yearmonth(dat$Date)
dat_ts <- as_tsibble(dat, index = Month)

fit <- model(
  dat_ts,
  var = VAR(vars(g1, g2) ~ AR(p = 1:6), ic = "bic")
)

report(fit)   

```

The two samples used are the MSA rates in Oxnard and LA. They both have 609 observations, which were taken monthly since 1975.

Looking at the above data (descriptive and forecasting), we can see that the growth rate in MSA1 (Oxnard) is 0.52%, very similar to that of MSA2 (LA), which is 0.53%. They two sets are very correlated (0.96). 

Using the fpp3 package's var() function, chosen using BIC to keep models simple, we find that a var(6) model leads to the most accurate results. BIC is minimized at p=6 which is why var(6) was chosen. In this model, almost all of the lag coefficients are significant, implying a strong correlation between prior lags and current lags. There are also clear ties between Oxnard rates and LA rates.


## 11.2

```{r 11.2}
p <- 6
n <- nrow(dat)
idx <- (p + 1):n

Y1 <- dat$g1[idx]  # Oxnard_t
Y2 <- dat$g2[idx]  # LA_t

X <- NULL
for (i in 1:p) {
  X <- cbind(X, dat$g1[idx - i], dat$g2[idx - i])
}
X <- as.data.frame(X)
names(X) <- unlist(lapply(1:p, function(i) c(paste0("g1_L", i), paste0("g2_L", i))))

df <- data.frame(Y1 = Y1, Y2 = Y2, X)

fit1_u <- lm(Y1 ~ ., data = df) 
fit2_u <- lm(Y2 ~ ., data = df)  

cols_g1 <- grep("^g1_L", names(df), value = TRUE)
cols_g2 <- grep("^g2_L", names(df), value = TRUE)

fit1_r <- lm(Y1 ~ ., data = df[, c("Y1", cols_g1), drop = FALSE]) 
fit2_r <- lm(Y2 ~ ., data = df[, c("Y2", cols_g2), drop = FALSE])  

gc_g2_to_g1 <- anova(fit1_r, fit1_u)  
gc_g1_to_g2 <- anova(fit2_r, fit2_u)  

gc_g2_to_g1
gc_g1_to_g2

```

Looking at the Granger Causality results, we see that regressing y2 on y1 and y1 on y2 both yeild significant p values (sub 5%). This means that LA Granger-causes Oxnard MSA and Oxnard Granger-causes LA MSA. In both cases, we can reject the "no cause" null hypothesis.


## 11.3

```{r 11.3}
h <- 36
future <- new_data(dat_ts, n = h)

irf_g1 <- IRF(fit, new_data = future, impulse = "g1")

irf_g2 <- IRF(fit, new_data = future, impulse = "g2")



autoplot(irf_g1, vars = vars(g1)) +
  labs(
    title = "Response of Oxnard (g1) to an Oxnard shock (g1)",
    x = "Months after shock",
    y = "Growth response (percentage points)"
  )

autoplot(irf_g1, vars = vars(g2)) +
  labs(
    title = "Response of LA (g2) to an Oxnard shock (g1)",
    x = "Months after shock",
    y = "Growth response (percentage points)"
  )

autoplot(irf_g2, vars = vars(g1)) +
  labs(
    title = "Response of Oxnard (g1) to an LA shock (g2)",
    x = "Months after shock",
    y = "Growth response (percentage points)"
  )

autoplot(irf_g2, vars = vars(g2)) +
  labs(
    title = "Response of LA (g2) to an LA shock (g2)",
    x = "Months after shock",
    y = "Growth response (percentage points)"
  )
```

The above graphs show that in forecasted data, self shocks (LA responding to LA shock, Oxnard responding to Oxnard shock) tend to be positive roughly 6 months later before a slight dip after a year. The counter shocks have an opposite effect, leading to a negative decline followed by an upturn once the self shocks begin to self correct.

Given my prior knowledge of the regions (growing up in LA and then moving to over the summer Oxnard), these results seem in line. I was surprised to see such similar growth rates since 1975, but the effects of shocks make sense to me. When LA grows, competing markets shrink, and vice versa.